---
title: "Designing your own misclassification model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing your own misclassification model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.width = 6, fig.height = 4)
library(misclassifyr)
```

The models this package ships with — record linkage, conditional
independence, fully non-parametric — are a starting menu, not a
description of your data. Occupational coding drift does not look like
name-matching failure. Age heaping does not look like either. If you use a
model whose shape is wrong, you will get a confident, precise, wrong
correction.

This vignette is about designing the right shape for your own setting. It
walks through four steps: enumerate the error channels, turn them into a
parameterized matrix, write that matrix as an R function, and verify it on
simulated data before it touches anything real. It then shows how to use a
large language model as a drafting assistant for step three, which is the
tedious one — and, at more length, how to check what the assistant gives
you. The division of labour is the whole point: **the assistant drafts, you
verify.** An assistant that produces a plausible-looking matrix with a
transposed index will cost you more time than writing it yourself, unless
you have a habit of checking.

## The object you are designing

$\Delta$ is the distribution of what you *observe* given what is *true*:

$$\Delta_{y^*,\,(y_1,y_2)} \;=\; \Pr(Y_1 = y_1,\, Y_2 = y_2 \mid Y^* = y^*).$$

Concretely, it is a `J` by `J^2` matrix, stored in the package as a plain
numeric vector of length `J^3`. Three facts about the layout are load-bearing
and are where hand-written (and generated) code goes wrong:

- **Rows are the truth.** Row `y*` of $\Delta$ is a probability distribution
  over the `J^2` possible pairs of recorded values. Rows sum to one.
- **Columns are pairs, with the first measure varying fastest.** Column
  `(y2 - 1) * J + y1` holds $\Pr(Y_1 = y_1, Y_2 = y_2 \mid Y^*)$.
- **The vector is column-major**, so `matrix(Delta, nrow = J)` recovers the
  matrix.

When the two measures err independently given the truth — the usual case,
and the assumption the package's built-in models make — you never build the
`J` by `J^2` object directly. You build two small `J` by `J` matrices,

$$\Delta^{(m)}_{y_m,\,y^*} = \Pr(Y_m = y_m \mid Y^* = y^*), \qquad m = 1, 2,$$

each with **columns** that sum to one, and combine them. Note the
transposition: in $\Delta^{(m)}$ the truth indexes columns; in $\Delta$ it
indexes rows. This one idiom does the combining, and it is worth copying
verbatim rather than re-deriving:

```{r combine}
combine_Delta <- function(Delta1, Delta2) {
  blocks <- lapply(1:nrow(Delta2), function(y2) diag(Delta2[y2, ]) %*% t(Delta1))
  c(do.call(cbind, blocks))
}
```

## Step 1: enumerate the error channels

Before writing any code, write down — in words — every way a value in your
outcome column could differ from the truth. In historical microdata the list
is usually short and drawn from four families.

**Transcription error.** A clerk, a keyer, or an OCR engine misreads the
source. Errors are *local in the written form*: "Baker" becomes "Barker",
1863 becomes 1868. They are not local in the analytic ordering — a
transcription error in an occupation string can land anywhere in the income
distribution. Shape: a diffuse matrix with a strong diagonal, with
off-diagonal mass concentrated on categories with similar spellings or
codes.

**Coding drift.** The same underlying job is coded differently in different
years, or by different agencies, or under a revised classification. Errors
are *systematic, not random*: every "agricultural labourer" in one source
becomes "farm hand" in another. Shape: mass concentrated in a few specific
off-diagonal cells, often asymmetric, and often stable across the
distribution of $X$. Crucially, this channel violates the "modal report is
the truth" condition if the drift is large enough to move the mode.

**Record-linkage failure.** The name-matching algorithm links the wrong two
people. The recorded outcome is then a stranger's outcome, drawn from
whatever pool of people the algorithm was choosing among. Shape: the
"slab" — a fraction $\alpha$ of each column replaced by a common
distribution $\rho$ that does not depend on the truth.

**Heaping.** Respondents and enumerators round. Ages pile on multiples of
five, incomes on round hundreds, birth years on decades. Shape: mass flows
*toward* focal categories and away from their neighbours. Heaping is the
channel most likely to surprise you, because it is not mean-reverting and
can bias a slope in either direction.

Two questions to ask about each channel you list:

1. *Does it depend on the truth?* Transcription and heaping do. Linkage
   failure, to a first approximation, does not — which is why it is the
   easiest to parameterize.
2. *Does it depend on the regressor?* If yes, you must condition on
   something that makes it not depend on the regressor, using the `controls`
   argument of `prep_misclassification_data()`. The model assumes errors are
   independent of $X$ given $Y^*$ (and given the controls).

## Step 2: turn channels into a parameterized $\Delta$

Each channel maps to a functional form with a handful of parameters. Three
forms cover most of what applied work needs.

### The record-linkage slab

$$\Delta^{(m)} = (1 - \alpha_m)\, I + \alpha_m\, \rho_m \mathbf{1}',$$

read as: with probability $1 - \alpha_m$ the link is right and you observe
the truth; with probability $\alpha_m$ it is wrong and you observe an
independent draw from $\rho_m$. Parameters: one rate $\alpha_m$ per measure,
plus the $J - 1$ free elements of $\rho_m$ if you estimate it rather than
plugging in a known pool distribution. This is `model_to_Delta_RL_ind()`,
and the factories `make_empirical_Delta_RL*()` build variants that plug in
the empirical margin for $\rho$.

Which $\rho$? Not, in general, the unconditional marginal distribution of
the outcome. A failed link draws from the pool of candidates the algorithm
was choosing among — people with similar names, similar birth years, and
*in the same place*, if the linker matched on place. Using the
unconditional margin when the true pool is narrower makes false links look
like correct ones and biases $\hat\alpha$ downward. The last section of this
vignette shows how to handle this properly.

### Monotone and ordinal forms

When the outcome is ordered — an occupational rank, an income decile, a
literacy scale — errors usually move you a short distance along the
ordering, not to an arbitrary category. The natural form is a band matrix:

$$\Delta^{(m)}_{y_m, y^*} = \begin{cases}
p^{\text{down}} & y_m = y^* - 1\\
1 - p^{\text{down}} - p^{\text{up}} & y_m = y^*\\
p^{\text{up}} & y_m = y^* + 1\\
0 & \text{otherwise,}
\end{cases}$$

with the boundary columns renormalized. Two parameters per measure describe
the entire matrix regardless of `J`, which is why ordinal designs stay
tractable when `J` is large. Richer variants let the rates vary with $y^*$,
or widen the band to two steps.

### Heaping forms

Fix a set of focal categories $F$ (multiples of five, round hundreds,
decades). With probability $h$ the reported value is snapped to the nearest
focal category; otherwise it is correct:

$$\Delta^{(m)}_{y_m, y^*} = (1 - h_m)\,\mathbf{1}\{y_m = y^*\} + h_m\,
\mathbf{1}\{y_m = f(y^*)\},$$

where $f(y^*)$ is the nearest element of $F$. One parameter per measure.
Note that this matrix has zero columns off the diagonal and the focal
column, and that it is diagonally dominant only while $h_m < 1/2$.

### Composing channels

Channels compose by matrix multiplication. If the truth is first
mis-transcribed ($T$) and the resulting record is then mis-linked ($L$),

$$\Delta^{(m)} = L \, T,$$

and because each factor has columns summing to one, so does the product.
Compose in the order the errors physically happened. Do not add channels —
adding two column-stochastic matrices does not give you one.

## Step 3: write the closure

`misclassifyr()` takes `model_to_Delta` as a **function of a single
unconstrained numeric vector** `psi`. The contract:

| Requirement | Why |
|---|---|
| Accepts one argument, `psi`, any real vector | The optimizer works in unconstrained space |
| Returns `c(Delta)`, a numeric vector of length `J^3` | The likelihood indexes it directly |
| Rows of `matrix(Delta, nrow = J)` sum to one | It is a conditional distribution |
| All entries non-negative | Same |
| `J` recoverable from `length(psi)`, or captured in the closure | The function is called with `psi` alone |

The last row is the reason these are written as *closures*: the design
function captures `J`, and any fixed inputs like $\rho$ or the focal set,
from its enclosing environment. Use a factory that takes the fixed inputs
and returns the closure.

Constrain parameters inside the closure, never outside. A probability comes
from `plogis(psi)`; a set of probabilities summing to one comes from a
softmax with a reference category. That way `psi` can be any real vector and
the optimizer never has to respect a boundary.

Here is a complete design — the record-linkage slab with a known pool
distribution $\rho$ and a common rate per measure:

```{r slab-design}
make_slab_design <- function(rho) {
  J <- length(rho)
  function(psi) {
    alpha <- plogis(psi)                       # two rates, in (0, 1)
    D1 <- diag(J) * (1 - alpha[1]) + outer(rho, rep(alpha[1], J))
    D2 <- diag(J) * (1 - alpha[2]) + outer(rho, rep(alpha[2], J))
    combine_Delta(D1, D2)
  }
}
```

Two practical notes.

**Always pass `psi_0`.** `misclassifyr()` knows starting values only for its
three built-in designs, which it recognizes by a `"name"` attribute. For a
custom design you must supply `psi_0` yourself. If you do not, and your
closure carries no `"name"` attribute, the function errors — which is the
outcome you want. If you have set a `"name"` attribute that the package does
not recognize *and* omitted `psi_0`, the starting value silently stays `NA`
and the optimization fails downstream with an obscure message. The simplest
rule: pass `psi_0`, and do not bother setting `"name"`.

**Know what the diagonal-dominance penalty does and does not cover.**
`misclassifyr()` adds a penalty, scaled by `lambda_dd`, whenever a row of
$\Delta$ fails to peak at the column corresponding to its own latent value
within some $Y_2$ block. Under conditional independence that is a constraint
on $\Delta^{(1)}$ only: each of its columns must have its largest entry on
the diagonal. $\Delta^{(2)}$ is left unconstrained, so if your second
measure could plausibly mis-report more often than it reports correctly, the
penalty will not stop it. The default `lambda_dd = sum(tab$n)^2` makes the
$\Delta^{(1)}$ constraint effectively hard, which is what rules out the
relabelled solution; lower it if your design legitimately needs to explore
that region.

### The matching log prior

If you want posterior inference (`bayesian = TRUE`) you also need
`log_prior_Delta`, a function of the same `psi`. It is *not* the log
density of your parameters; it is the log density of a chosen prior on the
**probabilities**, expressed in `psi` coordinates. Moving from probability
space to unconstrained space multiplies the density by the Jacobian
determinant of the link, so

$$\log p(\psi) = \log p\big(\Delta(\psi)\big) + \log\big|\det J_{\Delta}(\psi)\big|.$$

For a flat prior on a simplex reached by a softmax with a reference
category, `logit_link_volume()` returns exactly that term. For a single
probability reached by `plogis()`, the volume factor is `dlogis(psi, log =
TRUE)`. Two lines, and you get them right by checking them numerically
(below) rather than by rederiving them.

```{r slab-prior}
log_prior_slab <- function(psi) sum(dlogis(psi, log = TRUE))
```

## Step 4: verify before touching real data

Every design gets the same treatment: a structural check that costs
milliseconds, then a recovery check on simulated data. Write the harness
once and point it at everything.

```{r harness}
check_Delta_design <- function(f, psi, J, tol = 1e-8) {
  v <- f(psi)
  stopifnot("wrong length: expected J^3"      = length(v) == J^3)
  D <- matrix(v, nrow = J)
  stopifnot("wrong shape: expected J x J^2"   = all(dim(D) == c(J, J^2)),
            "negative entries in Delta"       = all(D >= -tol),
            "rows of Delta do not sum to one" = max(abs(rowSums(D) - 1)) < tol)

  # Diagonal dominance: within each Y2 block, row y* must peak at column y*
  peaks_ok <- all(vapply(1:J, function(y2) {
    blk <- D[, ((y2 - 1) * J + 1):(y2 * J), drop = FALSE]
    all(diag(blk) >= apply(blk, 1, max) - tol)
  }, logical(1)))

  # Every parameter must actually move the matrix. A block of psi that is
  # silently ignored is the single most common defect in generated code.
  moves <- vapply(seq_along(psi), function(i) {
    psi2 <- psi; psi2[i] <- psi2[i] + 0.25
    max(abs(f(psi2) - v)) > tol
  }, logical(1))

  list(structure_ok = TRUE,
       diagonally_dominant = peaks_ok,
       inert_parameters = which(!moves))
}
```

The last check is not paranoia. This package's own history includes a
released version in which the prior for the second misclassification matrix
was computed from the first matrix's block of `psi` — the second block was
inert, and nothing about the output looked wrong. Perturbing one coordinate
at a time catches it in a second.

Run it on the slab design:

```{r harness-run}
rho <- c(0.30, 0.25, 0.20, 0.15, 0.10)
slab <- make_slab_design(rho)
check_Delta_design(slab, psi = qlogis(c(0.2, 0.3)), J = 5)
```

### Recovery on simulated data

Structural correctness is necessary, not sufficient. The real test is
whether the estimator, fed data generated *from your design*, gets your
parameters back. Two helpers do this. For the built-in shapes,
`synthetic_data()` generates a tabulation directly:

```{r synth-builtin}
set.seed(1)
syn <- synthetic_data(J = 3, K = 3, I = 1, sample_size = 5000,
                      dgp_delta = "Record Linkage, independent, 10 - 30%")
head(syn$tab[[1]], 3)
```

For a custom design you simulate from it yourself. This is six lines, and it
is worth understanding because the cell ordering it produces — `X` fastest,
then `Y1`, then `Y2` — is exactly the ordering `misclassifyr()` expects:

```{r simulator}
simulate_tab <- function(Pi, Delta, N) {
  J <- nrow(Pi); K <- ncol(Pi)
  p <- c(t(Pi) %*% matrix(Delta, nrow = J))       # cell probabilities
  data.frame(X  = rep(1:K, J^2),
             Y1 = rep(rep(1:J, each = K), J),
             Y2 = rep(1:J, each = J * K),
             n  = as.numeric(rmultinom(1, N, p)))
}

# A diagonal-heavy latent joint distribution to simulate from
make_Pi <- function(J, K) {
  P <- outer(1:J, 1:K, function(j, k) exp(-0.5 * (j - k)^2))
  P / sum(P)
}
```

Now generate data with known $\alpha = (0.20, 0.30)$ and see whether the
estimator finds them:

```{r slab-recovery}
set.seed(13)
J <- 5; K <- 5
Pi_true    <- make_Pi(J, K)
alpha_true <- c(0.20, 0.30)
tab_slab   <- simulate_tab(Pi_true, slab(qlogis(alpha_true)), N = 2e5)

fit_slab <- misclassifyr(
  tab = tab_slab, J = J, K = K,
  X_names = as.character(1:K),
  Y1_names = as.character(1:J), Y2_names = as.character(1:J),
  model_to_Delta = slab, psi_0 = c(-1, -1),
  X_vals = 1:K, Y_vals = 1:J,
  mle = TRUE, bayesian = FALSE, makeplots = FALSE
)

round(plogis(tail(fit_slab$eta_hat_mle, 2)), 3)   # should be 0.20, 0.30
max(abs(matrix(fit_slab$Pi_hat_mle, nrow = J) - Pi_true))
```

If a design fails this test, it is either not identified or wrongly coded,
and no amount of real data will fix it.

## Using a language model as a design assistant

Step 3 — translating a shape you have already reasoned about into indexed R
code — is exactly the kind of task a large language model does well and you
do slowly. Steps 1, 2, and 4 are not. Delegate accordingly.

Four things make the difference between a useful draft and a plausible mess:

- **State the layout contract explicitly.** Do not assume the assistant
  knows this package. Paste the row/column convention every time.
- **Give the parameterization, do not ask for one.** You decided the shape
  in step 2; the assistant's job is to implement it, not to choose it.
- **Ask for the closure factory, not a bare function**, so that `J` and any
  fixed inputs are captured.
- **Ask it to state its assumptions.** A good response flags the ones you
  did not pin down — boundary handling, whether the rate varies with $y^*$
  — instead of silently choosing.

Below are four prompts of the kind that work, with what a good response
looks like and how to check it. In every case the check is the same harness
from step 4, and in every case you should run it before reading the code
closely.

### Prompt 1: the record-linkage slab with a known pool

````
I am using the R package misclassifyr. I need a `model_to_Delta` function.

The contract:
- The function takes one argument, `psi`, an unconstrained numeric vector.
- It returns `c(Delta)`, a numeric vector of length J^3 holding a J x J^2
  matrix column-major.
- Rows of that matrix index the latent outcome Y*; column (y2-1)*J + y1
  holds Pr(Y1 = y1, Y2 = y2 | Y* = y*). Rows sum to one.
- Errors in the two measures are conditionally independent given Y*, so
  build two J x J matrices Delta1, Delta2 with Pr(Ym = ym | Y* = y*) in
  entry [ym, y*] (columns sum to one) and combine them.

The model: each measure is a record-linkage slab. With probability
1 - alpha_m the recorded value equals Y*; with probability alpha_m it is an
independent draw from a KNOWN probability vector rho of length J. There are
exactly two free parameters, alpha_1 and alpha_2, constrained to (0,1)
inside the function.

Write it as a factory `make_slab_design(rho)` that returns the closure.
Also give me `psi_0`, a sensible starting value for a 15% error rate.
State any assumptions you had to make.
````

**What a good response looks like.** The factory captures `rho` and derives
`J <- length(rho)`; `alpha <- plogis(psi)` inside the closure; the slab built
as `diag(J) * (1 - alpha) + outer(rho, rep(alpha, J))`; the combination step
using the `diag(Delta2[y2, ]) %*% t(Delta1)` idiom; `psi_0 <-
rep(qlogis(0.15), 2)`. A good response also *tells* you that it assumed the
same `rho` for both measures and that `rho` is not renormalized after the
diagonal is added — both real choices you might have wanted differently.

**How to verify.** Run `check_Delta_design()`, and then — because this
design has a special case that the package already implements — check that
the two agree to machine precision. `model_to_Delta_RL_ind()` builds the
same slab, but parameterizes $\rho$ through a softmax rather than taking it
as given, so feeding it the `psi` that encodes our $\rho$ must reproduce our
matrix exactly:

```{r verify-1}
gen <- make_slab_design(rho)                     # the generated factory
alpha <- c(0.18, 0.24)

check_Delta_design(gen, qlogis(alpha), J = 5)

# The same slab, expressed in model_to_Delta_RL_ind's parameterization
r <- log(rho / rho[length(rho)])[1:(length(rho) - 1)]
psi_ref <- c(r, r, rep(qlogis(alpha[1]), 5), rep(qlogis(alpha[2]), 5))

max(abs(gen(qlogis(alpha)) - model_to_Delta_RL_ind(psi_ref)))
```

Cross-checking against an implementation you already trust is the strongest
verification available, and it is worth structuring your designs so that at
least one special case reduces to something shipped.

### Prompt 2: ordinal drift

````
Same package, same contract as before (J x J^2, rows are Y*, column
(y2-1)*J + y1, rows sum to one, conditional independence).

The model: the outcome is ORDERED (occupational rank 1..J). An error moves
the recorded value at most one rank. For measure m, define probabilities
(p_down_m, p_stay_m, p_up_m) summing to one, constant across ranks. At rank
1 there is no "down" and at rank J no "up": in those columns, drop the
impossible move and renormalize the remaining two probabilities.

Parameterize with a softmax using "stay" as the reference category, so psi
has 2 entries per measure (4 total) and is unconstrained.

Write it as `make_ordinal_design(J)` returning the closure. Give me psi_0
for roughly 10% down and 10% up. State your assumptions.
````

**What a good response looks like.** An inner helper that builds one `J` by
`J` matrix from `(p_down, p_up)`, a loop over columns that writes at most
three entries, and explicit renormalization in the two boundary columns.
`w <- exp(c(psi[1], 0, psi[2])); w <- w / sum(w)` for the softmax. The
response should flag that renormalizing at the boundary raises `p_stay`
there — a modelling choice, not a bug, but one you should have made
knowingly.

**How to verify.** Structural check, then recovery:

```{r ordinal}
make_ordinal_design <- function(J) {
  one_measure <- function(p_down, p_up) {
    D <- matrix(0, J, J)
    for (y in 1:J) {
      w <- c(down = if (y > 1) p_down else 0,
             stay = 1 - p_down - p_up,
             up   = if (y < J) p_up else 0)
      w <- w / sum(w)                             # renormalize at boundaries
      if (y > 1) D[y - 1, y] <- w[["down"]]
      D[y, y] <- w[["stay"]]
      if (y < J) D[y + 1, y] <- w[["up"]]
    }
    D
  }
  function(psi) {
    w1 <- exp(c(psi[1], 0, psi[2])); w1 <- w1 / sum(w1)
    w2 <- exp(c(psi[3], 0, psi[4])); w2 <- w2 / sum(w2)
    combine_Delta(one_measure(w1[1], w1[3]), one_measure(w2[1], w2[3]))
  }
}

J <- 4; K <- 4
ordinal <- make_ordinal_design(J)
psi_true <- c(log(0.10 / 0.80), log(0.10 / 0.80),   # measure 1: 10% / 10%
              log(0.05 / 0.80), log(0.15 / 0.80))   # measure 2:  5% / 15%

check_Delta_design(ordinal, psi_true, J = J)
```

Structure is fine and every parameter moves the matrix. Now recovery:

```{r ordinal-recovery}
set.seed(7)
Pi_true <- make_Pi(J, K)
tab_ord <- simulate_tab(Pi_true, ordinal(psi_true), N = 2e5)

fit_ord <- misclassifyr(
  tab = tab_ord, J = J, K = K,
  X_names = as.character(1:K),
  Y1_names = as.character(1:J), Y2_names = as.character(1:J),
  model_to_Delta = ordinal, psi_0 = rep(-2, 4),
  X_vals = 1:K, Y_vals = 1:J,
  mle = TRUE, bayesian = FALSE, makeplots = FALSE
)

psi_hat <- tail(fit_ord$eta_hat_mle, 4)
softmax <- function(x) { w <- exp(c(x[1], 0, x[2])); w / sum(w) }
rbind(truth     = c(softmax(psi_true[1:2]), softmax(psi_true[3:4])),
      estimated = c(softmax(psi_hat[1:2]),  softmax(psi_hat[3:4]))) |>
  round(3)
```

Both triples come back. And the point of the whole exercise — the
correction moves the coefficient in the right direction:

```{r ordinal-beta}
naive_beta <- local({
  P <- tapply(tab_ord$n, list(tab_ord$Y1, tab_ord$X), sum)
  Pi_to_beta_inner(c(P / sum(P)), 1:K, 1:J, 1)
})
c(truth     = Pi_to_beta_inner(c(Pi_true), 1:K, 1:J, 1),
  naive     = naive_beta,
  corrected = unname(Pi_to_beta(X_vals = 1:K, Y_vals = 1:J, mle = TRUE,
                                Pi_mle = fit_ord$Pi_hat_mle,
                                cov_Pi = fit_ord$cov_Pi_mle)$beta_hat_mle)) |>
  round(4)
```

### Prompt 3: heaping on focal values

````
Same package, same contract.

The model: reported values HEAP on a known set of focal categories. I will
pass `focal`, an integer vector of focal category indices. With probability
h_m the reported value is snapped to the nearest focal category (ties: the
smaller index); otherwise it is recorded correctly. One parameter per
measure, so psi has length 2, constrained to (0,1) with plogis().

Note this means a focal category is never mis-reported, and that a column
whose true value is already focal is degenerate. That is intended.

Write it as `make_heaping_design(J, focal)` returning the closure. Precompute
the nearest-focal lookup outside the closure so it is not recomputed on
every likelihood evaluation. State your assumptions.
````

**What a good response looks like.** `nearest <- sapply(1:J, function(y)
focal[which.min(abs(focal - y))])` computed **in the factory**, not the
closure — that detail matters, because the closure is called tens of
thousands of times inside `optim()`. The matrix built as `diag(J) * (1 - h)`
with `h` *added* to the `nearest[y]` entry of column `y`, so that a focal
column ends up degenerate at one rather than at `1 - h + h` split across two
cells. A good response warns that diagonal dominance requires `h < 0.5`.

**How to verify.** The structural harness catches the dominance issue
directly, so probe it at a value that should fail:

```{r heaping}
make_heaping_design <- function(J, focal) {
  nearest <- vapply(1:J, function(y) focal[which.min(abs(focal - y))], numeric(1))
  one_measure <- function(h) {
    D <- diag(J) * (1 - h)
    for (y in 1:J) D[nearest[y], y] <- D[nearest[y], y] + h
    D
  }
  function(psi) {
    h <- plogis(psi)
    combine_Delta(one_measure(h[1]), one_measure(h[2]))
  }
}

J <- 6; K <- 6
heaping <- make_heaping_design(J, focal = c(1, 3, 6))

check_Delta_design(heaping, qlogis(c(0.25, 0.40)), J = J)$diagonally_dominant
check_Delta_design(heaping, qlogis(c(0.65, 0.65)), J = J)$diagonally_dominant
```

The design is legitimate below one half and not above it, exactly as the
response should have warned. Recovery at an admissible value:

```{r heaping-recovery}
set.seed(11)
Pi_true <- make_Pi(J, K)
h_true  <- c(0.25, 0.40)
tab_hp  <- simulate_tab(Pi_true, heaping(qlogis(h_true)), N = 2e5)

fit_hp <- misclassifyr(
  tab = tab_hp, J = J, K = K,
  X_names = as.character(1:K),
  Y1_names = as.character(1:J), Y2_names = as.character(1:J),
  model_to_Delta = heaping, psi_0 = c(0, 0),
  X_vals = 1:K, Y_vals = 1:J,
  mle = TRUE, bayesian = FALSE, makeplots = FALSE
)

rbind(truth = h_true, estimated = plogis(tail(fit_hp$eta_hat_mle, 2))) |> round(3)
```

Both rates come back — and note the two warnings, which are *expected here
and should not alarm you*. A heaping matrix has structural zeros: a
non-focal column has mass on exactly two cells and nothing anywhere else.
That pushes parts of the estimated joint distribution against the boundary
of the simplex, where the likelihood cannot curve, so the information matrix
is singular by construction. The point estimates are unaffected; the
analytical standard errors are not usable. `vignette("inference")` explains
how to tell this benign case apart from a genuine identification failure,
and what to report instead.

And the substantive lesson that makes heaping worth modelling separately:

```{r heaping-beta}
naive_hp <- local({
  P <- tapply(tab_hp$n, list(tab_hp$Y1, tab_hp$X), sum)
  Pi_to_beta_inner(c(P / sum(P)), 1:K, 1:J, 1)
})
c(truth     = Pi_to_beta_inner(c(Pi_true), 1:K, 1:J, 1),
  naive     = naive_hp,
  corrected = unname(Pi_to_beta(X_vals = 1:K, Y_vals = 1:J, mle = TRUE,
                                Pi_mle = fit_hp$Pi_hat_mle,
                                cov_Pi = fit_hp$cov_Pi_mle)$beta_hat_mle)) |>
  round(4)
```

The naive estimate is *larger* than the truth. Misclassification does not
always attenuate: heaping pulls observations toward a small number of focal
categories, which here happen to be spread across the range, and the induced
error is correlated with the truth in a way that inflates the slope. Anyone
who assumed "measurement error biases towards zero, so my estimate is a
lower bound" would have drawn the wrong conclusion.

### Prompt 4: the matching log prior

````
Same package. I have this `model_to_Delta` closure:

    function(psi) {
      w1 <- exp(c(psi[1], 0, psi[2])); w1 <- w1 / sum(w1)
      w2 <- exp(c(psi[3], 0, psi[4])); w2 <- w2 / sum(w2)
      combine_Delta(one_measure(w1[1], w1[3]), one_measure(w2[1], w2[3]))
    }

I want `log_prior_Delta(psi)`: the log density, in psi coordinates, of a
FLAT prior on the underlying probability simplices. That means the log
absolute determinant of the Jacobian of the map from psi to probabilities,
summed over the two independent softmax blocks.

misclassifyr exports `logit_link_volume(x)` which returns exactly this
volume term for one softmax-with-reference block, up to an additive
constant that does not depend on x. Use it. Do not rederive the Jacobian.

Also tell me how I can check the answer numerically.
````

**What a good response looks like.** Three lines —
`logit_link_volume(psi[1:2]) + logit_link_volume(psi[3:4])` — plus, in
answer to the last sentence, a numerical check comparing
`logit_link_volume()` against `log(abs(det(numDeriv::jacobian(...))))` on
the free coordinates of the softmax. Be suspicious of any response that
computes the prior from `psi[1:2]` twice; that is the exact bug this package
once shipped, and it is an easy one to generate.

**How to verify.** Do the numerical check the response should have proposed:

```{r prior-check}
log_prior_ordinal <- function(psi) {
  logit_link_volume(psi[1:2]) + logit_link_volume(psi[3:4])
}

# The free coordinates of a softmax with a reference category
softmax_free <- function(x) { p <- exp(c(x, 0)); p <- p / sum(p); p[-length(p)] }

x <- c(-1.2, 0.4)
numeric_logdet <- log(abs(det(numDeriv::jacobian(softmax_free, x))))
c(numeric   = numeric_logdet,
  from_pkg  = logit_link_volume(x) - lfactorial(3))   # constant removed
```

They agree to machine precision. And the inertness check applies to priors
too:

```{r prior-inert}
base <- log_prior_ordinal(c(0, 0, 0, 0))
vapply(1:4, function(i) {
  psi <- c(0, 0, 0, 0); psi[i] <- 0.5
  abs(log_prior_ordinal(psi) - base) > 1e-10
}, logical(1))
```

All four `TRUE`: every coordinate reaches the prior.

## Stacking: enriching the outcome with what the linker ignored

There is one more design idea that is less about the shape of $\Delta$ than
about what you feed it, and it is often the highest-return change available.

Return to the record-linkage slab. A failed link draws a rival from the
pool of candidates the algorithm was choosing among. Which variables that
pool is restricted on turns out to matter twice over, in opposite-looking
ways that call for the same fix.

**If the algorithm ignored a variable, that variable is free information.**
Linking algorithms typically match on name, birth year, and birthplace, and
routinely ignore county or parish of *residence*, because people move.
That makes residence enormously informative about whether a link is real: a
correctly linked man is usually still somewhere plausible, whereas a rival
lands wherever the name pool happens to sit. Enriching the outcome with a
variable the linker never used — pairing occupation with county, say —
sharpens the contrast between correct and failed links well beyond what
either variable does alone. You are not adding a control; you are adding a
dimension along which the error is more visible.

**If the algorithm used a variable, ignoring that is a bias.** Suppose the
linker *did* match on birthplace. Then a rival is a same-birthplace
person, and a $\rho$ estimated from the unconditional distribution of
birthplace is too diffuse. Rivals that agree with the truth on birthplace
look like correct links, and $\hat\alpha$ is biased downward.

Both point to the same recipe: split the tabulation by the variable, and
estimate a **shared** error rate across the resulting cells. That is
`misclassifyr_rl_em_stacked()`. It takes a list of tabulations, holds
$\alpha_1, \alpha_2$ common across cells, and lets the rival distribution
$\rho_c$ and the joint distribution $\Pi_c$ vary freely within each — so the
error rate is estimated from all the data at once, while the rival pool is
allowed to be as local as it really is.

Here is the failure it prevents. Four birthplace cells; the linker matched
on birthplace, so a failed link draws a rival from the *same* cell:

```{r stacked}
set.seed(41)
C <- 4; J <- 20; N <- 8e4; alpha_true <- 0.25

cell <- sample.int(C, N, replace = TRUE)
base <- (cell - 1) * 5                                  # 5 counties per cell
xi   <- base + sample.int(5, N, replace = TRUE)         # father's county
j    <- ifelse(runif(N) < 0.8, xi,
               base + sample.int(5, N, replace = TRUE)) # son's true county
y1   <- ifelse(runif(N) < alpha_true,
               base + sample.int(5, N, replace = TRUE), j)
y2   <- ifelse(runif(N) < alpha_true,
               base + sample.int(5, N, replace = TRUE), j)
df   <- data.frame(cell = cell, X = xi, Y1 = y1, Y2 = y2)

tabulate_cells <- function(d) {
  aggregate(list(n = rep(1, nrow(d))),
            by = list(X = d$X, Y1 = d$Y1, Y2 = d$Y2), FUN = sum)
}

pooled  <- misclassifyr_rl_em(tabulate_cells(df), J = J, K = J)
stacked <- misclassifyr_rl_em_stacked(lapply(split(df, df$cell), tabulate_cells),
                                      J = J, K = J)

rbind(pooled  = pooled$alpha,
      stacked = stacked$alpha) |> round(3)
```

The pooled estimator, which implicitly treats the rival pool as the whole
country, understates the false-link rate by about a quarter of its value.
The stacked estimator recovers it. If you then use $\hat\alpha$ to correct a
downstream coefficient, that error propagates directly into the correction.

The practical recipe:

1. Find a variable that the linking algorithm did **not** use, or used
   differently from how you will condition on it. County or parish of
   residence is the usual candidate in historical data; linkers routinely
   ignore it because people move.
2. Tabulate separately within each value of that variable.
3. Estimate with `misclassifyr_rl_em_stacked()`, which shares $\alpha$ and
   lets everything else vary by cell.
4. Check that $\hat\alpha$ is stable as you coarsen or refine the cells. If
   it drifts a lot, the conditioning variable is doing something other than
   what you think.

The per-cell output carries everything you need for the downstream step:

```{r stacked-out}
str(stacked$cells[[1]], max.level = 1)
```

`vignette("sparse-and-large")` picks up from here: how to run this when the
conditioning variable has hundreds of values and the outcome thousands, and
how to hold $\hat\alpha$ fixed from a coarse first stage while estimating a
fine-grained $\Pi$ in a second.
