---
title: "Getting started: correcting an estimate for linkage error"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: correcting an estimate for linkage error}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.width = 6, fig.height = 4)
library(misclassifyr)
```

## The problem

You have linked historical records: fathers to sons, censuses to censuses,
draft cards to death certificates. You link them with a name-matching
algorithm, you get a linked sample, and you run a regression on it.

Some of those links are wrong. Not many, perhaps — 10%, 20% — but wrong. And
a wrong link does not put a random number in your outcome column; it puts
*somebody else's* outcome there. Somebody else drawn, roughly, from the pool
of people whose names looked similar.

The consequence for your regression is not subtle. Suppose you regress a
son's income on his father's income to estimate intergenerational
persistence. When a son is mis-linked, his recorded income is really a
stranger's income, which is uncorrelated with *this* father's income. The
slope you estimate is a mixture of the true slope (for correctly linked
pairs) and zero (for the rest). You get attenuation, and the size of the
attenuation is roughly the false-link rate. A 20% link error rate knocks 20%
off your estimate of social mobility's opposite.

The usual fixes are unsatisfying. You can tighten the linking algorithm,
which trades link errors for selection bias. You can bound the damage under
worst-case assumptions, which is honest but wide. Or you can *model* the
error, estimate its distribution from the data, and correct the estimate.
That is what this package does.

## What you need: two noisy measures

The one non-negotiable requirement is that the outcome be measured **twice**.

Here is why, without any econometrics. With a single measure you cannot tell
a genuinely poor son from a rich son whose record was mis-linked to a poor
stranger. Both appear as "poor" in your data, and nothing in the data
separates them. Add a second measure of the same underlying outcome — the
son observed again in a later census, say — and the separation becomes
possible. When the two measures agree, the value is probably right. When
they disagree, at least one is wrong. *How often* they disagree, and *in
which direction*, is exactly the information you need to pin down the error
rates.

Once you know the error rates, you can undo them: the distribution you
actually observe is the true distribution passed through a known
"smearing" matrix, and you invert the smearing.

Three conditions make this work.

1. **Two measures.** Call them `Y1` and `Y2`. They must be measures of the
   same latent outcome `Y*`.
2. **Conditionally independent errors.** Given the truth, the two measures
   must err independently. Two transcriptions by the same clerk from the
   same page are *not* two measures — they share their mistakes. Two
   different censuses, or a linked record and an independently collected
   self-report, usually are.
3. **Errors unrelated to the regressor.** Given the truth, the errors must
   not depend on `X`. If the linking algorithm is worse for poor families
   than rich ones, condition on family status using the `controls` argument
   so that this holds within each control cell.

There is a fourth condition that is more of a normalization: the modal
report is the truth. Most sons are recorded correctly. Without this the
model cannot tell which of two candidate solutions is the "true" outcome and
which is the error, and the package would happily hand you the relabeled
one.

## The data

The package ships a synthetic dataset, `ancienregime`: 100,000 father–son
pairs in Third Estate France, linked across three censuses (that never
happened) in 1750, 1770, and 1780. The son's occupation and income score are
recorded **twice**, in 1770 and again in 1780, which supplies the two
measures.

```{r data}
data(ancienregime)
str(ancienregime)
```

The two measures of the son disagree for a substantial share of pairs:

```{r disagree}
mean(ancienregime$son_occupation_1770 != ancienregime$son_occupation_1780)
```

That disagreement is not a nuisance. It is the identifying variation.

## Step 1: tabulate

`misclassifyr` works from a tabulation, not from microdata. The estimator
only ever touches the counts of each `(X, Y1, Y2)` cell, so
`prep_misclassification_data()` builds that table and records the
bookkeeping the estimator needs — how many categories, what they are called,
and what numeric score each category carries.

Occupation is categorical, but we want a regression coefficient in income
units. The `_bin` arguments handle exactly this: bin the continuous income
variable by occupation, and record the within-bin mean income as the numeric
score for each category.

```{r prep}
inputs <- prep_misclassification_data(
  data          = ancienregime,
  outcome_1     = "son_income_1780",
  outcome_2     = "son_income_1770",
  regressor     = "father_income_1750",
  outcome_1_bin = "son_occupation_1780",
  outcome_2_bin = "son_occupation_1770",
  regressor_bin = "father_occupation_1750",
  weights       = "linked_weight",
  record_vals   = TRUE,
  round_vals    = 0
)

c(J = inputs$J, K = inputs$K)     # categories of the outcome and the regressor
nrow(inputs$tab)                  # a balanced J^2 * K table
head(inputs$tab)
round(inputs$Y_vals)              # mean income within each occupation
```

`J` is the number of outcome categories, `K` the number of regressor
categories, and the tabulation has `J^2 * K = 216` rows: one for every
combination of the father's occupation and the two recorded occupations of
the son.

## Step 2: estimate the misclassification model

`misclassifyr()` maximizes the likelihood of that tabulation over two
objects at once:

- $\Pi$, the joint distribution of the regressor $X$ and the **latent**
  outcome $Y^*$ — the thing you actually want; and
- $\Delta$, the distribution of the two recorded values given the truth,
  $\Pr(Y_1, Y_2 \mid Y^*)$ — the smearing matrix.

You choose the shape of $\Delta$ through the `model_to_Delta` argument. Here
we use `model_to_Delta_RL_ind`, the record-linkage model: with probability
$1 - \alpha$ the link is correct and the recorded occupation is the true
one; with probability $\alpha$ the link failed and the recorded occupation
is an independent draw from a distribution $\rho$ over occupations. That is
a natural description of name-matching failure, and it costs only $4J - 2$
parameters rather than the $2J(J-1)$ of an unrestricted model.

```{r fit}
set.seed(1)   # the stability check below perturbs starting values
fit <- misclassifyr(
  tab            = inputs$tab,
  J              = inputs$J,
  K              = inputs$K,
  X_names        = inputs$X_names,
  Y1_names       = inputs$Y1_names,
  Y2_names       = inputs$Y2_names,
  X_vals         = inputs$X_vals,
  Y_vals         = inputs$Y_vals,
  model_to_Delta = model_to_Delta_RL_ind,
  mle            = TRUE,
  bayesian       = FALSE,
  makeplots      = FALSE
)
```

## Step 3: read the output

`misclassifyr()` returns a long list. Four pieces matter most.

**The estimated joint distribution.** `Pi_hat_mle` is a length-`J*K` vector
holding the `J` by `K` matrix column-major: rows are the latent outcome,
columns the regressor.

The categories are labelled by their mean income, since that is what we
binned on; they are the six occupations, in ascending order of income.

```{r pi}
occ <- c("Vagabond", "Metayer", "Journalier",
         "Petit Metiers", "Petite Bourgeoisie", "Haute Bourgeoisie")

Pi_hat <- matrix(fit$Pi_hat_mle, nrow = inputs$J,
                 dimnames = list(`Son (latent)` = occ, Father = occ))
round(Pi_hat, 3)
```

**The estimated error rates.** `Delta_hat_mle` holds the `J` by `J^2` joint
distribution of both measures given the truth. To read off the more
intuitive object — $\Pr(Y_1 \mid Y^*)$, the chance a son whose true
occupation is $j$ is recorded as $k$ in 1780 — marginalize over the second
measure:

```{r delta}
marginalize <- do.call(rbind, replicate(inputs$J, diag(inputs$J), simplify = FALSE))
Delta1 <- matrix(fit$Delta_hat_mle, nrow = inputs$J) %*% marginalize
dimnames(Delta1) <- list(Truth = occ, `Recorded 1780` = occ)
round(Delta1, 3)
```

Careful with the orientation: **rows are the truth**, so row $j$ is the
distribution of what gets *recorded* for a son whose occupation really is
$j$. The diagonal is the probability of being recorded correctly, and one
minus the diagonal is the implied error rate:

```{r alpha}
round(1 - diag(Delta1), 3)   # implied error rate by true occupation
```

These are worth staring at. They are not uniform — the model says the
lowest-status category is recorded correctly far less often than the
middling ones, which is what you would expect if vagabonds are hard to
find, hard to link, and easy to confuse with day labourers. If instead they
had come back implausible (60% error where the linkage literature reports
10%, or a rate of zero) that would be evidence against the shape of
$\Delta$ you chose, not a result.

**Two diagnostics you should always look at.**

```{r diag}
fit$fisher_info_err      # was the information matrix invertible?

c(elements_of_theta = inputs$J * inputs$K + inputs$J^3,
  inconsistency     = fit$inconsistency_mle)
```

The first tells you whether analytical standard errors are trustworthy;
here they are.

The second re-runs the optimizer from a perturbed starting value and reports
the *total* absolute distance between the two answers, summed over every
element of $\Pi$ and $\Delta$. `misclassifyr()` warns above 0.1, which is a
blunt threshold: it does not scale with the size of the model, and this
model has 252 elements. Per element the discrepancy is around 0.003, small
enough to be convergence slop rather than a second optimum. Still, a fired
warning is a prompt to check whether the number you care about is actually
sensitive to it, and `vignette("inference")` shows how to tell — and what to
do when it genuinely is.

## Step 4: turn $\Pi$ into a regression coefficient

`Pi_to_beta()` computes the slope of the population least-squares projection
of $Y^*$ on $X$ implied by the estimated joint distribution, together with a
delta-method standard error.

```{r beta}
corrected <- Pi_to_beta(
  X_vals  = inputs$X_vals,
  Y_vals  = inputs$Y_vals,
  mle     = TRUE,
  Pi_mle  = fit$Pi_hat_mle,
  cov_Pi  = fit$cov_Pi_mle
)

c(beta = unname(corrected$beta_hat_mle),
  se   = unname(corrected$se_beta_mle))
```

## Does the correction bite?

Compare against what you would have reported without it. The cleanest
comparison holds the discretization fixed and simply *skips* the correction:
take the observed joint distribution of the 1780 measure and the father's
occupation and compute the same slope from it.

```{r naive}
tab <- inputs$tab

naive_from <- function(measure) {
  P <- tapply(tab$n, list(tab[[measure]], tab$X), sum)
  Pi_to_beta_inner(c(P / sum(P)), inputs$X_vals, inputs$Y_vals, 1)
}

results <- data.frame(
  estimate = c("Naive, 1780 measure",
               "Naive, 1770 measure",
               "Individual-level OLS",
               "Misclassification-corrected"),
  beta = c(naive_from("Y1"),
           naive_from("Y2"),
           unname(coef(lm(son_income_1780 ~ father_income_1750,
                          data = ancienregime,
                          weights = ancienregime$linked_weight))[2]),
           unname(corrected$beta_hat_mle))
)
results$ratio_to_corrected <- results$beta / unname(corrected$beta_hat_mle)
knitr::kable(results, digits = 3)
```

The naive estimates come in 13% to 19% below the corrected one, which is the
order of magnitude the estimated error rates above imply. Reported as an
intergenerational elasticity in a paper, that gap is the difference between
a society you would describe as fairly mobile and one you would describe as
notably less so — from the same data, with the same linking algorithm.

Note also that the two naive estimates disagree with each other (0.59 versus
0.63). They are estimates of the same quantity from the same sample; the gap
between them is measurement error making itself visible. When your results
move depending on which of two equivalent measures you happen to use, that
is a signal to model the error rather than pick a favourite.

## Where to go next

- **`vignette("designing-misclassification-models")`** — the important one.
  The record-linkage model used above is one shape of $\Delta$ among many,
  and it will not describe *your* errors. This vignette shows how to
  enumerate the error channels in your own setting, turn them into a
  parameterized $\Delta$, write it as an R closure, and verify it. It also
  covers using a large language model as a drafting assistant, and how to
  check what it hands you.
- **`vignette("sparse-and-large")`** — when the outcome has hundreds or
  thousands of categories (counties, detailed occupation codes) the balanced
  tabulation above becomes impossible and the general estimator too slow.
  Sparse tabulations and a dedicated expectation-maximization estimator
  handle it.
- **`vignette("inference")`** — delta-method standard errors, what a
  singular information matrix means, and Monte Carlo confidence sets that
  remain valid when the model is only partially identified.
