---
title: "High-dimensional outcomes: sparse tabulations and the EM estimator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{High-dimensional outcomes: sparse tabulations and the EM estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.width = 6, fig.height = 4)
library(misclassifyr)
```

## Why the balanced table stops working

`misclassifyr()` works from a **balanced** tabulation: one row for every
combination of the regressor and the two recorded outcomes, including the
combinations that never occur. That is `J^2 * K` rows.

For a six-category occupation scheme that is 216 rows and you never think
about it. For anything at the granularity historical work actually cares
about, the arithmetic turns hostile:

```{r arithmetic}
sizes <- data.frame(
  outcome = c("occupation, 6 groups", "US state", "occupation, 3-digit codes",
              "US county", "parish / township"),
  J = c(6, 50, 300, 3000, 15000)
)
sizes$K <- sizes$J
sizes$rows_balanced <- format(sizes$J^2 * sizes$K, big.mark = ",",
                              scientific = FALSE)
knitr::kable(sizes)
```

A county-level analysis would need a table with twenty-seven billion rows.
That is not a table you build and then decide is too slow; it is a table
that does not fit in memory on any machine you have.

But almost all of those rows are zero. In a linked sample of half a million
records spread over 3,000 counties, at most 500,000 cells can be non-empty,
out of twenty-seven billion. Two things follow, and this vignette is about
both:

1. **Zero-count cells contribute nothing to the likelihood.** So you can
   drop them, if you keep track of where the surviving cells sit. That is
   `prep_misclassification_data(sparse = TRUE)`.
2. **The general estimator is still the wrong tool.** Even with a sparse
   table, `misclassifyr()` optimizes over a parameter vector containing all
   of $\Pi$ — `J * K - 1` free parameters — using a quasi-Newton method with
   numerically differentiated gradients. At `J = 300` that is 89,999
   parameters. For the record-linkage model specifically, there is a much
   better route: an expectation-maximization algorithm with closed-form
   updates, which is `misclassifyr_rl_em()`.

## Sparse tabulations

`sparse = TRUE` keeps only the cells with positive counts and adds a
`cell_idx` column recording each surviving cell's position in the balanced
layout. `misclassifyr()` accepts either form and returns identical
estimates — the sparse path exists purely to keep the tabulation from
exploding.

```{r sparse-prep}
data(ancienregime)
occ <- c("Vagabond", "Metayer", "Journalier",
         "Petit Metiers", "Petite Bourgeoisie", "Haute Bourgeoisie")

args <- list(
  data      = ancienregime,
  outcome_1 = "son_occupation_1780",
  outcome_2 = "son_occupation_1770",
  regressor = "father_occupation_1750",
  X_names   = occ, Y1_names = occ, Y2_names = occ,
  weights   = "linked_weight"
)

dense  <- do.call(prep_misclassification_data, args)
sparse <- do.call(prep_misclassification_data, c(args, list(sparse = TRUE)))

c(dense_rows = nrow(dense$tab), sparse_rows = nrow(sparse$tab))
head(sparse$tab, 3)
```

Here the saving is trivial — one empty cell out of 216 — because this
dataset has only six categories. The point is what `cell_idx` buys you: it
lets the likelihood look up each observed cell's model probability without
ever materializing the full grid.

```{r sparse-equiv}
identical(dense$tab$n[sparse$tab$cell_idx], sparse$tab$n)
```

That equality is the entire contract. Everything downstream is unchanged.

## The EM estimator for the record-linkage model

For the record-linkage model — each measure equals the truth with
probability $1 - \alpha_m$ and is otherwise an independent draw from
$\rho_m$ — the likelihood of an observed cell factorizes into exactly four
terms, one per configuration of the two links:

| Configuration | Cell probability contribution |
|---|---|
| Both links correct | $(1-\alpha_1)(1-\alpha_2)\,\Pi_{y_1, x}\,\mathbf{1}\{y_1 = y_2\}$ |
| First correct, second failed | $(1-\alpha_1)\,\alpha_2\,\rho_{2,y_2}\,\Pi_{y_1, x}$ |
| First failed, second correct | $\alpha_1\,\rho_{1,y_1}\,(1-\alpha_2)\,\Pi_{y_2, x}$ |
| Both failed | $\alpha_1\alpha_2\,\rho_{1,y_1}\rho_{2,y_2}\,\Pi_{+, x}$ |

Every M-step update is then a closed-form weighted average: $\alpha_m$ is
the expected share of failed links, $\rho_m$ the expected distribution of
failed-link draws, $\Pi$ the expected latent-cell counts. No numerical
optimization, no Hessian, no $J$ by $J^2$ matrix. Each iteration costs
O(observed cells), and neither the misclassification matrices nor the
balanced tabulation is ever formed.

The support of $\Pi$ is fixed at initialization to the union of the observed
$(Y_1, X)$ and $(Y_2, X)$ pairs. EM keeps zero cells at zero, so this is not
a restriction so much as a recognition of one.

### A `J = 300` example

Simulate a linked sample over 300 "counties": with probability 0.2 the son's
true county differs from his father's, and each of the two recorded counties
is a failed link with probability $\alpha_m$.

```{r em-dgp}
rl_dgp <- function(J, N, alpha1, alpha2, stay = 0.8, seed = 23) {
  set.seed(seed)
  xi <- sample.int(J, N, replace = TRUE)                 # father's county
  moves <- sample.int(3, N, replace = TRUE)
  j <- ifelse(runif(N) < stay, xi, ((xi - 1 + moves) %% J) + 1)
  rho <- tabulate(j, J) / N
  y1 <- ifelse(runif(N) < alpha1, sample.int(J, N, replace = TRUE, prob = rho), j)
  y2 <- ifelse(runif(N) < alpha2, sample.int(J, N, replace = TRUE, prob = rho), j)
  tab <- aggregate(list(n = rep(1, N)),
                   by = list(X = xi, Y1 = y1, Y2 = y2), FUN = sum)
  list(tab = tab, rho = rho,
       Pi_true = unclass(table(factor(j, 1:J), factor(xi, 1:J))) / N)
}

d <- rl_dgp(J = 300, N = 4e5, alpha1 = 0.20, alpha2 = 0.35)

c(balanced_rows = 300^2 * 300, observed_cells = nrow(d$tab))
round(nrow(d$tab) / (300^2 * 300), 4)                 # fill rate
```

Under one percent of the balanced grid is ever visited, and that is with a
sample four hundred thousand records deep. Now estimate.

```{r em-fit}
t0 <- Sys.time()
fit <- misclassifyr_rl_em(d$tab, J = 300, K = 300)
elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

c(alpha1 = unname(fit$alpha["alpha1"]), alpha2 = unname(fit$alpha["alpha2"]),
  iterations = fit$n_iter, seconds = round(elapsed, 1))
```

Both rates recovered, in a couple of seconds, on a problem the general
estimator cannot represent. The log likelihood increases monotonically, as
EM guarantees — a useful sanity check on any run:

```{r em-monotone}
all(diff(fit$loglik_trace) > -1e-6)
```

The estimated joint distribution comes back on its support only, as a long
data frame of `(j, i, p)` triples:

```{r em-pi}
head(fit$Pi, 3)
nrow(fit$Pi)

# Accuracy against the truth, on the support
p_true <- d$Pi_true[cbind(fit$Pi$j, fit$Pi$i)]
c(correlation = cor(fit$Pi$p, p_true),
  max_abs_err = max(abs(fit$Pi$p - p_true)))
```

Also worth checking: $\hat\rho$, the phantom distribution, against the true
latent margin.

```{r em-rho}
max(abs(fit$rho1 - d$rho))
```

## Two-step estimation with `alpha_fixed`

Estimating $\alpha$ and a 90,000-cell $\Pi$ from the same data is
uncomfortable. The number of parameters grows with the sample, which is the
classic incidental-parameters setting: $\hat\alpha$ inherits bias from the
noise in $\hat\Pi$, and its sampling distribution is not the textbook one.

The fix is to split the two jobs across two levels of aggregation.

1. **Coarse first step.** Aggregate the outcome to a level where cells are
   dense and the asymptotics are standard — states rather than counties,
   one-digit rather than three-digit occupation codes. Estimate $\alpha$
   there. A false link is a false link regardless of how finely you code
   the outcome, so this $\hat\alpha$ transfers.
2. **Fine second step.** Estimate $\Pi$ at the granularity you actually
   want, with $\alpha$ held fixed at the first-step value.

`alpha_fixed` implements step two:

```{r two-step}
# Step 1: aggregate 300 counties into 30 "states" and estimate alpha there
to_state <- function(v) ((v - 1) %/% 10) + 1
coarse <- aggregate(list(n = d$tab$n),
                    by = list(X  = to_state(d$tab$X),
                              Y1 = to_state(d$tab$Y1),
                              Y2 = to_state(d$tab$Y2)),
                    FUN = sum)

step1 <- misclassifyr_rl_em(coarse, J = 30, K = 30)

rbind(truth  = c(alpha1 = 0.20, alpha2 = 0.35),
      coarse = step1$alpha,
      fine   = fit$alpha) |> round(3)
```

The coarse step lands on the truth, and slightly closer than the fine-grained
free-$\alpha$ run — which is the incidental-parameters point in miniature.
Now hold it fixed and re-estimate $\Pi$ at full granularity:

```{r two-step-2}
step2 <- misclassifyr_rl_em(d$tab, J = 300, K = 300,
                            alpha_fixed = unname(step1$alpha))

step2$alpha                                   # held, not estimated
p2 <- d$Pi_true[cbind(step2$Pi$j, step2$Pi$i)]
c(free_alpha  = max(abs(fit$Pi$p   - p_true)),
  fixed_alpha = max(abs(step2$Pi$p - p2)))
```

One caveat on step one. Aggregation is only harmless if a failed link rarely
lands inside the same coarse cell as the truth. That holds here, because the
phantom pool is spread over all 300 counties and only one in thirty draws
falls in the right state. It fails if the linking algorithm itself matched on
something correlated with the aggregation — if it matched on region, a
phantom is a same-region person, aggregation to regions hides most link
failures, and $\hat\alpha$ from the coarse step is biased toward zero. In
that case condition on the matching variable instead and share $\alpha$
across cells with `misclassifyr_rl_em_stacked()`; see
`vignette("designing-misclassification-models")`.

You can also fix $\rho_1$ and $\rho_2$ the same way, through the `rho1` and
`rho2` arguments, when you know the candidate pool the linking algorithm was
drawing from — for instance because you can reconstruct it from the index
the linker searched.

## Separating true transitions from linkage error with `T2`

Here is a problem that only appears at high dimension. Suppose the outcome
is county of residence and the two measures are ten years apart. The son's
county in the second observation may differ from the first for two entirely
different reasons: the link failed, or *he moved*. The plain record-linkage
model cannot tell these apart, so it charges genuine migration to
$\alpha_2$ and reports a false-link rate that is too high.

The `T2` argument fixes this by giving the second measure its own
true-transition operator:

$$\Delta^{(2)} = (1 - \alpha_2)\, T + \alpha_2\, \mathbf{1}\rho_2',$$

read as: a *correct* link observes a draw from $T$ applied to the latent
value — real migration between the two dates — while a failed link still
draws from $\rho_2$. `T` is supplied as a sparse data frame with columns
`j` (latent category), `l` (observed category) and `t` (probability), rows
summing to one within `j`. It is a plug-in, held fixed and not estimated,
which is the point: you discipline it with external information — a census
migration question, published inter-county flows, a gravity model fitted to
aggregate data — rather than asking the same data to identify both.

```{r t2-dgp}
set.seed(31)
J <- 50; N <- 1.5e5; a1 <- 0.10; a2 <- 0.15

xi <- sample.int(J, N, replace = TRUE)
jj <- ifelse(runif(N) < 0.8, xi, ((xi + sample.int(4, N, replace = TRUE) - 1) %% J) + 1)
rho <- tabulate(jj, J) / N

# The son really moves between the two observation dates, 15% of the time
moved  <- sample.int(3, N, replace = TRUE)
l_true <- ifelse(runif(N) < 0.85, jj, ((jj - 1 + moved) %% J) + 1)

y1 <- ifelse(runif(N) < a1, sample.int(J, N, replace = TRUE, prob = rho), jj)
y2 <- ifelse(runif(N) < a2, sample.int(J, N, replace = TRUE, prob = rho), l_true)
tab <- aggregate(list(n = rep(1, N)), by = list(X = xi, Y1 = y1, Y2 = y2), FUN = sum)
```

The known migration structure, as a `T2` data frame:

```{r t2-operator}
T2 <- do.call(rbind, lapply(1:J, function(j) {
  data.frame(j = j,
             l = c(j, ((j - 1 + 1:3) %% J) + 1),
             t = c(0.85, rep(0.05, 3)))
}))
head(T2, 4)
```

Estimating with and without it:

```{r t2-fit}
with_T2    <- misclassifyr_rl_em(tab, J = J, K = J, T2 = T2)
without_T2 <- misclassifyr_rl_em(tab, J = J, K = J)

rbind(truth      = c(alpha1 = a1, alpha2 = a2),
      with_T2    = with_T2$alpha,
      without_T2 = without_T2$alpha) |> round(3)
```

Ignoring migration nearly doubles the estimated false-link rate for the
second measure, while leaving the first — which is contemporaneous with the
regressor and so unaffected — alone. If you then used that inflated
$\hat\alpha_2$ to correct a coefficient, you would over-correct by the size
of the migration rate.

## Choosing between the estimators

| | `misclassifyr()` | `misclassifyr_rl_em()` |
|---|---|---|
| Shape of $\Delta$ | any design you write | record-linkage slab only |
| Feasible `J` | tens | thousands |
| Tabulation | balanced, or sparse with `cell_idx` | observed cells only |
| Inference | delta method, posterior draws, Monte Carlo sets | point estimates; bootstrap or the two-step architecture |
| Controls | `controls` argument, estimated in parallel | `misclassifyr_rl_em_stacked()`, shared $\alpha$ |
| Extras | diagonal-dominance penalty, priors, trace plots | `alpha_fixed`, `rho1`/`rho2`, `T2` |

The practical rule: if `J` is small enough that `J^2 * K` rows is a table
you would be willing to print the dimensions of, use `misclassifyr()` and
get the full inferential menu. If it is not, and your errors are linkage
failures, use the EM estimator. If your errors at high dimension are *not*
linkage failures, you are in genuinely hard territory and should think
about aggregating the outcome before you think about estimators.

See `vignette("inference")` for what to report from either, and
`vignette("designing-misclassification-models")` for how to choose the shape
of $\Delta$ in the first place.
